---
title: "Discovering Join Paths with map_join_paths()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discovering Join Paths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(data.table)
library(DBmaps)
```

## Introduction
In the previous vignettes, we learned how to define rich metadata for our tables using table_info() and how to use that metadata to automatically generate aggregation code. 
The next logical question is: how do we know which tables can be joined together?

This vignette introduces the map_join_paths() function. It acts as a discovery tool that analyzes your complete metadata and automatically identifies all possible direct join connections between your tables. 
The output is a "Join Map" that serves as the foundational "road map" of your database for all subsequent join operations.

## Create the Master Metadata
The map_join_paths() function requires the master_metadata_dt as its input. 
Let's create the metadata for our four retail data tables (customers, products, transactions, and views) to build a complete example.

```{r master-metadata}
# Define metadata for all four tables
customers_info <- table_info(
  table_name = "customers",
  source_identifier = "customers.csv",
  identifier_columns = "customer_id",
  key_outcome_specs = list(
    list(OutcomeName = "CustomerCount", ValueExpression = 1, AggregationMethods = list(
      list(AggregatedName = "CountByRegion", AggregationFunction = "sum", GroupingVariables = "region")
    ))
))

products_info <- table_info(
  table_name = "products",
  source_identifier = "products.csv",
  identifier_columns = "product_id",
  key_outcome_specs = list(
    list(OutcomeName = "ProductCount", ValueExpression = 1, AggregationMethods = list(
      list(AggregatedName = "ProductsPerCategory", AggregationFunction = "sum", GroupingVariables = "category")
    ))
))

transactions_info <- table_info(
  table_name = "transactions",
  source_identifier = "transactions.csv",
  identifier_columns = c("customer_id", "product_id", "time"),
  key_outcome_specs = list(
    list(OutcomeName = "Revenue", ValueExpression = quote(price * quantity), AggregationMethods = list(
      list(AggregatedName = "RevenueByCustomer", AggregationFunction = "sum", GroupingVariables = "customer_id"),
      list(AggregatedName = "RevenueByProduct", AggregationFunction = "sum", GroupingVariables = "product_id")
    ))
  )
)

views_info <- table_info(
  table_name = "views",
  source_identifier = "views.csv",
  identifier_columns = c("customer_id", "product_id", "view_timestamp"),
  key_outcome_specs = list(
    list(OutcomeName = "ViewCount", ValueExpression = 1, AggregationMethods = list(
        list(AggregatedName = "ViewsByProduct", AggregationFunction = "sum", GroupingVariables = "product_id"),
        list(AggregatedName = "ViewsByCustomer", AggregationFunction = "sum", GroupingVariables = "customer_id")
    ))
  )
)

# Combine into a single master metadata table
master_metadata_dt <- rbindlist(list(customers_info, products_info, transactions_info, views_info))
cat("Master metadata table created successfully.\n")
```

## Generate the Join Map
Now that we have our master_metadata_dt, we can simply pass it to map_join_paths() to discover all the direct connections.

```{r join-the-map}
# Generate the Join Map from the metadata
join_map <- map_join_paths(master_metadata_dt)

# Print the resulting map
print(join_map)
```

## Understanding the Output
The join_map data.table is the "road map" of our database. Each row represents one possible direct join. The columns are:
table_from: The name of the table that would be aggregated (typically a "fact" table).
table_to: The name of the table it can be joined to (typically a "dimension" table).
join_key: The common column that connects the two tables.

Let's interpret the first row:
1:   transactions      customers   customer_id

This means: "An aggregated version of the transactions table, when grouped by customer_id, can be directly joined to the customers table using the customer_id column."
The function successfully found this path because customer_id is a grouping_variable for transactions and the identifier_column for customers.

## Conclusion
The map_join_paths() function automates the discovery of relational connections within your data schema. 
It transforms your abstract metadata definitions into a concrete map of potential joins.